Easy as Pi: A Network Coding Raspberry Pi Testbed

### 2016-06-06 testbed.tex
finished resizing image

% Define variable]
\begin{lstlisting}[]
$ export IMAGE="2016-05-27-raspbian-jessie-lite"
$ export WORKDIR="${HOME}/Raspbian"
\end{lstlisting}
\FloatBarrier

@@

% Root example
\begin{lstlisting}[]
$ whoami
<USERNAME>
$ sudo whoami
root
\end{lstlisting}
\FloatBarrier

@@

% Download and unpack image
\begin{lstlisting}[]
$ mkdir -p ${WORKDIR}
$ cd ${WORKDIR}
$ wget http://kom.aau.dk/project/TuneSCode/raspi/${IMAGE}.zip
$ unzip ${IMAGE}.zip
\end{lstlisting}
\FloatBarrier

@@

\subsection{(Optional) Resize image}
1.2G is not very much to hold the root filesystem given that it is desired to
in later steps alter the files and install more packages. It may therefore be
neccessary to increase the partition size slightly.
The following steps will therefore illustrate how image and root filesystem
can be expanded by one \ac{GB}.

First, expand the image one \ac{GB}:
\begin{lstlisting}[]
# dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img
\end{lstlisting}
\FloatBarrier

@@

@@

To browser and alter the files in the image, it is possible to mount the partitions.

Lets first mount the root partition. This is done by first creating a new folder
that we will call root which will be located in our working directory. After this,
we mount the partition starting at sector 137216. To provide this offset information
to mount, we need to find the offset in bytes. From the fdisk command we could also
see that each sector is 512 bytes. Thus, we multiply the sector size and the sector
start to get the offset in bytes.

\begin{lstlisting}[]
$ export ROOTDIR="${WORKDIR}/root"
$ mkdir -p ${ROOTDIR}
# mount -o loop,offset=$((137216*512)) ${IMAGE}.img ${ROOTDIR}
\end{lstlisting}
\FloatBarrier

@@

Mount boot partition:
\begin{lstlisting}[]
# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier

@@

% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/nodes.csv}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
# Ethernet MAC    Hostname
b8:27:eb:5b:da:20 rasp00
b8:27:eb:7b:c3:91 rasp01
b8:27:eb:54:9c:64 rasp02
b8:27:eb:95:bd:11 rasp03
\end{lstlisting}
\FloatBarrier

@@

# Assign hostname found in nodes.csv
if [ ! -z $hostname ]; then
    echo $hostname > /etc/hostname
    hostname $hostname
fi
\end{lstlisting}
\FloatBarrier

@@

% Get the files
\begin{lstlisting}[]
$ wget http://kom.aau.dk/project/TuneSCode/raspi/rasp_config.zip
# unzip rasp_config.zip -d ${ROOTDIR}/home/pi/
\end{lstlisting}
\FloatBarrier

"nodes.csv" can be changed using a text editor. For vim, this could be done
the following way:
\begin{lstlisting}[]
# vim ${ROOTDIR}/home/pi/rasp_config/nodes.csv
\end{lstlisting}
\FloatBarrier

@@

\begin{lstlisting}[]
bash /home/pi/rasp_config/set_hostname"
\end{lstlisting}
\FloatBarrier

@@

% CHROOT to OS image
\begin{lstlisting}[]
# apt-get install binfmt-support qemu qemu-user-static
# update-binfmts --display qemu-arm
qemu-arm (enabled):
     package = qemu-user-static
        type = magic
      offset = 0
       magic = \x7fELF\x01\x01\x01\x00\x00\x00\x00\x00\x00\x00\x00\x00\x02\x00\x28\x00
        mask = \xff\xff\xff\xff\xff\xff\xff\x00\xff\xff\xff\xff\xff\xff\xff\xff\xfe\xff\xff\xff
 interpreter = /usr/bin/qemu-arm-static
    detector = 
\end{lstlisting}
\FloatBarrier
% This was neccessary on arch
%# update-binfmts --importdir /var/lib/binfmts/ --import

@@

\begin{lstlisting}[]
# update-binfmts --enable qemu-arm
\end{lstlisting}
\FloatBarrier

@@

% CHROOT to OS image
\begin{lstlisting}[]
# cd $ROOTDIR
# cp /etc/resolv.conf etc/resolv.conf
\end{lstlisting}
\FloatBarrier

@@

% CHROOT to OS image
\begin{lstlisting}[]
# cp /usr/bin/qemu-arm-static usr/bin
\end{lstlisting}
\FloatBarrier

@@

% CHROOT to OS image
\begin{lstlisting}[]
# mount -t proc proc proc/
# mount --rbind /sys sys/
# mount --rbind /dev dev/
\end{lstlisting}
\FloatBarrier

@@

% CHROOT to OS image
\begin{lstlisting}[]
# proot -q qemu-arm-static -S ${ROOTDIR}
\end{lstlisting}
\FloatBarrier

@@

% CHROOT to OS image
\begin{lstlisting}[]
# chroot ${ROOTDIR} /usr/bin/qemu-arm-static /bin/bash    (ALTERNATIVE TO proot)
\end{lstlisting}
\FloatBarrier

@@

% Optionally, we may create a unique prompt to indicate we have changed root
\begin{lstlisting}[]
# export PS1="(chroot) $PS1"
\end{lstlisting}
\FloatBarrier

@@

% Update system
\begin{lstlisting}[]
# apt-get update
# apt-get upgrade
\end{lstlisting}
\FloatBarrier

Lets install some useful applications:
% Install packages
\begin{lstlisting}[]
# apt-get install vim git screen
\end{lstlisting}
\FloatBarrier

@@

% Get files
\begin{lstlisting}[]
# OVERLAYROOTDIR="/tmp/overlayroot"
# git clone https://github.com/chesty/overlayroot.git $OVERLAYROOTDIR
\end{lstlisting}
\FloatBarrier

@@

% Install required packages
\begin{lstlisting}[]
# apt-get install busybox
\end{lstlisting}
\FloatBarrier

@@

% Setup initramfs
\begin{lstlisting}[]
# cp ${OVERLAYROOTDIR}/hooks-overlay /etc/initramfs-tools/hooks/
# cp ${OVERLAYROOTDIR}/init-bottom-overlay /etc/initramfs-tools/scripts/init-bottom/
# echo "overlay" > /etc/initramfs-tools/modules
\end{lstlisting}
\FloatBarrier

@@

% Tell uboot to load initramfs
\begin{lstlisting}[]
# echo "initramfs init.gz" > /boot/config.txt
\end{lstlisting}
\FloatBarrier

@@

% Create an initramfs
\begin{lstlisting}[]
# ls /lib/modules/
\end{lstlisting}
\FloatBarrier

@@

% Create an initramfs
\begin{lstlisting}[]
# mkinitramfs -o /boot/init.gz -k 4.4.11+
\end{lstlisting}
\FloatBarrier

@@

% Exit chroot, umount, and write to memory card
\begin{lstlisting}[]
# exit
# cd ..
# umount --recursive ${ROOTDIR}
# losetup -d WHICH_NUMBE.img
\end{lstlisting}
\FloatBarrier

@@

\begin{lstlisting}[]
# lsblk
# df -h
\end{lstlisting}
\FloatBarrier

For us, the device was /dev/mmcblk0, so we write the image to that device:
\begin{lstlisting}[]
# dd if=${IMAGE}.img of=/dev/mmcblk0 bs=4M
\end{lstlisting}
\FloatBarrier

>>>>

% Define variable]
\begin{lstlisting}[]
$ export IMAGE="2016-05-27-raspbian-jessie-lite"
$ export WORKDIR="${HOME}/Raspbian"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Root example
\begin{lstlisting}[]
$ whoami
<USERNAME>
$ sudo whoami
root
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Download and unpack image
\begin{lstlisting}[]
$ mkdir -p ${WORKDIR}
$ cd ${WORKDIR}
$ wget http://kom.aau.dk/project/TuneSCode/raspi/${IMAGE}.zip
$ unzip ${IMAGE}.zip
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

\subsection{Resize image}
1.2G is not very much to hold the root filesystem given that it is desired to
in later steps alter the files and install more packages. It may therefore be
neccessary to increase the partition size slightly.
The following steps will therefore illustrate how image and root filesystem
can be expanded by one \ac{GB}.

First, expand the image one \ac{GB} and wait for data to be completely written
the disk:
\begin{lstlisting}[]
# dd if=/dev/zero bs=1M count=1024 >> ${IMAGE}.img && sync
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

Device                               Boot  Start     End Sectors  Size Id Type
2016-05-27-raspbian-jessie-lite.img1        8192  137215  129024   63M  c W95 FAT32 (LBA)
2016-05-27-raspbian-jessie-lite.img2      137216 2709503 2572288  1.2G 83 Linux
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Now, to expand the root filesystem it is required to first remove the Linux partition
and then add it again one \ac{GB} bigger.
We use fdisk for the task. The echo's within the parenthesis will function as
key-presses in \texttt{fdisk} and hence delete last partition, create a new
primary partition, print new partition table, and write new partition table
to the image:

\begin{lstlisting}[]
# (echo d; echo ; echo n; echo p; echo ; echo 137216; echo ; echo p; echo w) | fdisk ${IMAGE}.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Make the Raspbian image available as a block device: 
\begin{lstlisting}[]
# DEV=$(losetup --show -f -P ${IMAGE}.img); echo $DEV
/dev/loop0
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

We see that the image is now available as /dev/loop0. Use \texttt{lsblk}
to view the partitions of our image:

\begin{lstlisting}[]
# lsblk
NAME      MAJ:MIN RM  SIZE RO TYPE MOUNTPOINT
...
loop0       7:0    0  2.3G  0 loop 
|-loop0p1 259:2    0   63M  0 loop 
|-loop0p2 259:3    0  2.2G  0 loop
...
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Check and resize the root filesystem:
\begin{lstlisting}[]
# e2fsck -f ${DEV}p2
# resize2fs ${DEV}p2
\end{lstlisting}
\FloatBarrier

@@

Browsing and altering the files in the image is easy. Simply mount the partitions.
%To browser and alter the files in the image, it is possible to mount the partitions.

Lets first mount the root partition. This is done by first creating a new folder
that we will name \texttt{root} which will be located in our working directory.
After this, simply mount the partitions:
%After this,
%we mount the partition starting at sector 137216. To provide this offset information
%to mount, we need to find the offset in bytes. From the fdisk command we could also
%see that each sector is 512 bytes. Thus, we multiply the sector size and the sector
%start to get the offset in bytes.

\begin{lstlisting}[]
$ export ROOTDIR="${WORKDIR}/root"
$ mkdir -p ${ROOTDIR}
# mount ${DEV}p2 ${ROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((137216*512)) ${IMAGE}.img ${ROOTDIR}

@@

Mount boot partition:
\begin{lstlisting}[]
# mount ${DEV}p1 ${ROOTDIR}/boot
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
%# mount -o loop,offset=$((8192*512)) ${IMAGE}.img ${ROOTDIR}/boot

@@

% MAC and Hostname file
\Suppressnumber\begin{lstlisting}[]
<@\textcolor{gray}{\$\{ROOTDIR\}/home/pi/rasp\_config/nodes.csv}@>
<@\textcolor{gray}{
---------------------------------------------------------------}
\Reactivatenumber @>
# Ethernet MAC    Hostname
b8:27:eb:5b:da:20 rasp00
b8:27:eb:7b:c3:91 rasp01
b8:27:eb:54:9c:64 rasp02
b8:27:eb:95:bd:11 rasp03
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

# Assign hostname found in nodes.csv
if [ ! -z $hostname ]; then
    echo $hostname > /etc/hostname
    hostname $hostname
fi
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Get the files
\begin{lstlisting}[]
$ wget http://kom.aau.dk/project/TuneSCode/raspi/rasp_config.zip
# unzip rasp_config.zip -d ${ROOTDIR}/home/pi/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

"nodes.csv" can be changed using a text editor. For vim, this could be done
the following way:
\begin{lstlisting}[]
# vim ${ROOTDIR}/home/pi/rasp_config/nodes.csv
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

\begin{lstlisting}[]
bash /home/pi/rasp_config/set_hostname"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% CHROOT to OS image
\begin{lstlisting}[]
# apt-get install binfmt-support qemu qemu-user-static
# update-binfmts --display qemu-arm
qemu-arm (enabled):
     package = qemu-user-static
        type = magic
      offset = 0
       magic = \x7fELF\x01\x01\x01\x00\x00\x00\x00\x00\x00\x00\x00\x00\x02\x00\x28\x00
        mask = \xff\xff\xff\xff\xff\xff\xff\x00\xff\xff\xff\xff\xff\xff\xff\xff\xfe\xff\xff\xff
 interpreter = /usr/bin/qemu-arm-static
    detector = 
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
% This was neccessary on arch
%# update-binfmts --importdir /var/lib/binfmts/ --import

@@

\begin{lstlisting}[]
# update-binfmts --enable qemu-arm
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% CHROOT to OS image
\begin{lstlisting}[]
# cd $ROOTDIR
# cp /etc/resolv.conf etc/resolv.conf
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% CHROOT to OS image
\begin{lstlisting}[]
# cp /usr/bin/qemu-arm-static usr/bin
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% CHROOT to OS image
\begin{lstlisting}[]
# mount -t proc proc proc/
# mount --rbind /sys sys/
# mount --rbind /dev dev/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% CHROOT to OS image
\begin{lstlisting}[]
# proot -q qemu-arm-static -S ${ROOTDIR}
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% CHROOT to OS image
\begin{lstlisting}[]
# chroot ${ROOTDIR} /usr/bin/qemu-arm-static /bin/bash    (ALTERNATIVE TO proot)
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Optionally, we may create a unique prompt to indicate we have changed root
\begin{lstlisting}[]
# export PS1="(chroot) $PS1"
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Update system
\begin{lstlisting}[]
# apt-get update
# apt-get upgrade
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

Lets install some useful applications:
% Install packages
\begin{lstlisting}[]
# apt-get install vim git screen
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Get files
\begin{lstlisting}[]
# OVERLAYROOTDIR="/tmp/overlayroot"
# git clone https://github.com/chesty/overlayroot.git $OVERLAYROOTDIR
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Install required packages
\begin{lstlisting}[]
# apt-get install busybox
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Setup initramfs
\begin{lstlisting}[]
# cp ${OVERLAYROOTDIR}/hooks-overlay /etc/initramfs-tools/hooks/
# cp ${OVERLAYROOTDIR}/init-bottom-overlay /etc/initramfs-tools/scripts/init-bottom/
# echo "overlay" > /etc/initramfs-tools/modules
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Tell uboot to load initramfs
\begin{lstlisting}[]
# echo "initramfs init.gz" > /boot/config.txt
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Create an initramfs
\begin{lstlisting}[]
# ls /lib/modules/
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Create an initramfs
\begin{lstlisting}[]
# mkinitramfs -o /boot/init.gz -k 4.4.11+
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

% Exit chroot, umount, and write to memory card
\begin{lstlisting}[]
# exit
# cd ..
# umount --recursive ${ROOTDIR}
# losetup -d WHICH_NUMBE.img
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

@@

\begin{lstlisting}[]
# lsblk
# df -h
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}

For us, the device was /dev/mmcblk0, so we write the image to that device:
\begin{lstlisting}[]
# dd if=${IMAGE}.img of=/dev/mmcblk0 bs=4M
\end{lstlisting}
\FloatBarrier
\vspace{-5mm}
